Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2015-08-10 article.Rmd
Adds to intro, conclusion

It is important to understand our method’s limitations as well as its advantages. The approach of inferring gender from personal names is still a blunt way to study a complex subject. Gender theorists and feminist scholars have spent decades unpacking the full range of meanings behind gender as an analytical concept. On its own, our method has little to say about crucial topics such as the intersectionality of gender with race, class, and power, or the fluid relationship between biological sex and individual identity [Butler 1990, Butler 1993, Crenshaw 1991, McCall 2005]. The tool itself does not reflect how an individual person identifies oneself and is neutral as to what the categories “male” or “female” actually mean in any given historical and social context. The question of gender identity is especially relevant for our method, given its reliance on large datasets created by state agencies such as the U.S. Census Bureau. Much like the state’s definition of racial categories, governments impose their own vision of a gender binary onto their subjects, which in turn exerts a tremendous definitional power over societies and individuals [Brubaker and Cooper 2000, 14-17, Pascoe 2009, Canaday 2011, Landrum 2014]. Researchers who use this method need to be fully aware of the ways it replicates this problematic, state-defined gender binary.

As with any tool, the onus is on the researcher to understand how its underlying data was gathered, what it implies, and how it reflects social and cultural practices. Imperfect as it may be, this method nevertheless gives digital humanities scholars an important means of studying gender in textual collections. We have made the method as transparent and flexible as possible through the inclusion of probability figures - ex. 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. This makes the underlying data visible to users and allows them the opportunity to interrogate the assumptions that come with it. When used thoughtfully, this method can infer additional information about individuals that stand at the heart of humanities scholarship, providing another way to see and interpret sources across large scales of analysis.

@@

Include in conclusion
- gender is a tool that humanities scholars can put all of their tremendous skills in contextualizing, reocgnizing subtlety, and putting it in conversation with other sturctural power of race class and sexuality.
- Point to examples of how much it's downloaded, Ted Underwood, and especially Ben Schmidt's student evaluations, which is exactly the kind of ways it can be used to shed light on hidden patterns of systemic bias and discrimination.
- Posner's latest blog post on impreative of creating humanistic tools

Our algorithm for predicting gender from first names, when combined with historical datasets in the Gender package, provides a powerful new method for researchers interested in the study of gender. The digital turn has made available very large datasets, yet in the midst of abundance many humanists continue to deal with the problem of incompleteness and scarcity within their sources. Our method uses *abundant* existing datasets to infer additional information from *scarce* datasets. The temporal dimension of our work matches historical questions to a historical method, providing much improved results over simple, anachronistic, or ahistorical lookup methods. This method is also extensible to different times and places, provided that a researcher has a suitable reference dataset of names and genders from the period or area under study. The Gender package reverses the traditional disciplinary direction between the humanities and computer science. Rather than adapting existing computational methodologies, our method uses a humanistic starting point---the historical relationship between gender and naming practices---to contribute a new method to computer science.

@@

@@

[The Poetess Archive n.d.] The Poetess Archive. "About the Poetess Archive." <http://idhmc.tamu.edu/poetess/about/index.html>. Accessed 23 June 2015.]

@@

@@

[Weingart 2014] Weingart, Scott. "The Moral Role of DH in a Data-Driven World." *the scottbot irregular*. 9 September 2014, <http://www.scottbot.net/HIAL/?p=40944>

>>>>

It is important to understand our method’s limitations as well as its advantages. The approach of inferring gender from personal names is still a blunt way to study a complex subject. Gender theorists and feminist scholars have spent decades unpacking the full range of meanings behind gender as an analytical concept, including the fluid relationship between biological sex and individual identity [Butler 1990, Butler 1993]. On its own, our method has little to say about crucial topics such as the intersectionality of gender with race, class, and power, or what the categories “male” or “female” actually mean in any given historical and social context [Crenshaw 1991, McCall 2005]. Perhaps most importantly, the tool itself does not reflect how an individual person self-identifies and how gender shapes their lived experience. The question of gender identity is especially relevant for our method, given its reliance on large datasets created by state agencies such as the U.S. Census Bureau. Much like the state’s definition of racial categories, governments impose their own vision of a gender binary onto their subjects in ways that are deeply problematic. This gender binary reflects existing power structures more than the lived identities of individuals. The state's gender categorization exerts a tremendous definitional sway over societies and the people who comprise them, most notably in its systematic exclusion of transgender identities. [Brubaker and Cooper 2000, 14-17, Pascoe 2009, Canaday 2011, Landrum 2014].

As with any tool, researchers who use this method need to be fully aware of the ways it replicates a problematic, state-defined gender binary. The onus is on the user to understand how its underlying data was gathered, what it implies, and how it reflects social and cultural practices. Imperfect as it may be, this method nevertheless gives digital humanities scholars an important means of studying gender in textual collections. We have made the method as transparent and flexible as possible through the inclusion of probability figures - ex. 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. This makes the underlying data visible to users and allows them the opportunity to interrogate the assumptions that come with it. When used thoughtfully, this method can infer additional information about individuals who stand at the heart of humanities scholarship, providing another way to see and interpret sources across large scales of analysis.

@@

The digital turn has made available very large datasets, yet in the midst of abundance many humanists continue to deal with the problem of incompleteness and scarcity within their sources. Our method uses *abundant* existing datasets to infer additional information from *scarce* datasets. Its major contribution is to incorporate a historical approach to this algorithm, one that takes into account how gendered naming practices have changed over time. This temporal methodology provides much improved results over simpler, anachronistic, or ahistorical lookup methods. The Gender package is, of course, far from perfect. By relying on historical datasets collected by state agencies, it reinforces a gender binary and larger power structure that obscures how people may identify themselves and how they experience the world. To borrow a phrase from Miriam Posner, the Gender package does not meet the "radical, unrealized potential of digital humanities." But it might.

The Gender package is entirely open-source, and we have provided detailed and transarent documentation of its code and data. Researchers who wish to build upon it can modify the package accordingly. To take one imagined scenario: Facebook allows its users to either select from a large spectrum of gender identities or to write in their own prefered identity. Because these identities are associated with the user's first name and date of birth, this is exactly the kind of historical dataset that could be incorporated into the Gender package. Doing so would replace the package's simplistic categories of "Male" and "Female" with ones like "Female to Male", "Bigender", or "Cisgender Female". Although this imagined scenario would create its own set of challenges, it would represent a major improvement in how the Gender package dealt with the complexity and fluidity of gender identity. ^[This imagined scenario runs into its own set of problems. Facebook's business model makes it unlikely to grant researchers open access to its proprietary and closely-guarded datasets. And even if it did grant access to this data, Facebook collected it for particular purposes and in particular ways. Just as government agencies like the Social Security Administration collect data to serve the purposes of the American state, Facebook's data collection reflects the priorities of a global corporation rather than the lived experience of its users.] By building as much flexibility and transparency into the method as possible, we hope that others can continue to make just these kinds of improvements.

Even as it currently stands, the Gender package meets a pressing methodological need. Since its initial release in February 2014, the Gender package has been downloaded more than ________ times [??????] and incorporated as a pedagogical tool into digital humanities courses [Jockers 2014]. Researchers have used it to uncover gender inequity in a range of different areas, from biases in students' teaching evaluations to disparities between attendees and authors at the annual Digital Humanities Conference [Schmidt 2015, Weingart 2015]. These implementations help illustrate the value of the method. Its major intervention is to algorithmically inferring gender in textual sources in a way that takes into account change over time. Rather than trying to shoehorn existing computational methods, the Gender package uses a humanistic starting point---the historical relationship between gender and naming practices---to contribute a new method to computer science. This reverses the traditional disciplinary direction between the humanities and computer science.

@@

[Jockers 2014] Jockers, Matthew. "Experimenting with “gender” package in R." *Matthew L. Jockers*. 25 February 2015, <http://www.matthewjockers.net/2014/02/25/experimenting-with-gender-package-in-r/>. Accessed 10 August 2015.

@@

[The Poetess Archive n.d.] The Poetess Archive. "About the Poetess Archive." <http://idhmc.tamu.edu/poetess/about/index.html>. Accessed 23 June 2015.

[Posner 2015] Posner, Miriam. "What’s Next: The Radical, Unrealized Potential of Digital Humanities." *Miriam Posner's Blog*. 27 July 2015, <http://miriamposner.com/blog/whats-next-the-radical-unrealized-potential-of-digital-humanities/>. Accessed 10 August 2015.

@@

[Schmidt 2015] Schmidt, Ben. "Gender and Teacher Reviews." February 2015. <http://benschmidt.org/profGender/>.

@@

[Weingart 2014] Weingart, Scott. "The Moral Role of DH in a Data-Driven World." *the scottbot irregular*. 9 September 2014, <http://www.scottbot.net/HIAL/?p=40944>.

[Weingart 2015] Weingart, Scott. "Acceptances to Digital Humanities 2015 (part 3)." *the scottbot irregular*. 27 June 2015, <http://www.scottbot.net/HIAL/?p=41355>.
